---
title: "ECD Connect: What Usage Data Reveals About Practitioner Adoption"
author: "Kim Tichmann · Data & Product Strategy, DGMT"
date: "2025"
format:
  html:
    theme: cosmo
    css: styles.css
    toc: false
    smooth-scroll: true
    code-fold: true
    code-tools: false
    embed-resources: true
execute:
  echo: false
  message: false
  warning: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width  = 10,
  fig.height = 6,
  dpi        = 150,
  out.width  = "100%"
)

library(tidyverse)
library(scales)
```

<div class="page-header">
<h1>ECD Connect: What Usage Data Reveals About Practitioner Adoption</h1>
<p class="page-byline">Kim Tichmann · Data & Product Strategy Lead - ECD Connect, DGMT</p>
</div>

---

## Context: The problem this platform is trying to solve

Early childhood education (ECD) in South Africa is under-resourced and fragmented. Most of the sector's ~40,000 preschools operate informally, with no digital records and no way to systematically track whether children are attending or developing well.

ECD Connect is a live digital platform giving ECD practitioners tools to track attendance, monitor child development, manage finances, and access teaching resources. It currently serves 3,000+ registered users across South Africa.

This analysis examines what the platform's usage data reveals about how, and whether, practitioners actually adopt digital tools in a low-resource context. Where do they drop off? Which features drive sustained use? And what does that tell us about how to design and prioritise features for this population?

<div class="insight-box teal">
<p class="insight-label">Why this matters beyond ECD</p>
<p>Digital adoption patterns in the ECD sector mirror broader challenges in public-sector and emerging-market tech: how do you design for users who are time-poor, cautious about new technology, and operating without reliable infrastructure? The signals here have direct relevance to any product targeting underserved communities at scale.</p>
</div>
---

## Platform scale

Data covers ECD Connect App (open access) and ECD Connect Partner (white-label) tenants. ECD Connect Partner launched in October 2024 and ECD Connect App launched in April 2025. All data covers from launch to *19 May 2026*.

<div class="stat-block">
  <div class="stat-card teal"><p class="stat-number">3,276</p><p class="stat-label">Registered users</p></div>
  <div class="stat-card green"><p class="stat-number">1,384</p><p class="stat-label">Converted users (valid preschool)</p></div>
  <div class="stat-card yellow"><p class="stat-number">977</p><p class="stat-label">Preschools added</p></div>
  <div class="stat-card pink"><p class="stat-number">4141</p><p class="stat-label">Children registered</p></div>
</div>

**Conversion rate: 42%** of registered users complete onboarding to the point of a valid preschool connection. Understanding why the remaining 59% don't, and what distinguishes those who do, is the focus of this analysis.

---

## Part 1: The onboarding funnel

### How we define the funnel

To get meaningful value from ECD Connect, a practitioner must complete a seven-step activation sequence: registering, adding profile details, connecting to a preschool, and ultimately taking attendance. Each step requires deliberate action and represents a commitment to using the platform.

Note that a small subset of users engage regularly without completing the full funnel, likely using the app for resources or lesson planning rather than administrative tracking. The funnel below focuses on the core activation pathway.

The funnel below shows cumulative completion rates (OA and WL tenants).

```{r funnel-chart}
# ── Load data ─────────────────────────────────────────────────────────────────
# Expected CSV columns: Step, Count, `% of registered`
df_funnel <- read_csv("~/Desktop/data/data-usage/funnel-counts.csv")

df_funnel <- df_funnel |>
  mutate(
    step = case_when(
      grepl("Registered",  Step, ignore.case = TRUE) ~ "Step 0: Registered",
      grepl("Added name",  Step, ignore.case = TRUE) ~ "Step 1: Added name",
      grepl("ID number",   Step, ignore.case = TRUE) ~ "Step 2: Added ID number",
      grepl("preschool",   Step, ignore.case = TRUE) ~ "Step 3: Connected to preschool",
      grepl("class",       Step, ignore.case = TRUE) ~ "Step 4: Has a class",
      grepl("child",       Step, ignore.case = TRUE) ~ "Step 5: Added a child",
      grepl("attendance",  Step, ignore.case = TRUE) ~ "Step 6: Took attendance",
      TRUE ~ Step
    ),
    step         = fct_inorder(step),
    pct_of_total = as.numeric(`% of registered`),
    count        = as.numeric(Count)
  )

# ── Palette ───────────────────────────────────────────────────────────────────
n           <- nrow(df_funnel)
bar_colours <- colorRampPalette(c("#1D9E75", "#EF9F27", "#E24B4A"))(n)

df_funnel <- df_funnel |>
  mutate(
    bar_fill  = rev(bar_colours),
    r         = col2rgb(bar_fill)[1,] / 255,
    g         = col2rgb(bar_fill)[2,] / 255,
    b         = col2rgb(bar_fill)[3,] / 255,
    luminance = 0.2126 * r + 0.7152 * g + 0.0722 * b,
    text_col  = if_else(luminance > 0.45, "#2C2C2A", "white")
  )

# ── Plot ──────────────────────────────────────────────────────────────────────
ggplot(df_funnel, aes(x = fct_rev(step), y = count)) +

  geom_col(aes(fill = step), width = 0.65) +
  scale_fill_manual(values = setNames(df_funnel$bar_fill, df_funnel$step), guide = "none") +

  geom_text(
    aes(label = comma(count), colour = step),
    hjust    = 1.15,
    fontface = "bold",
    size     = 3.5
  ) +
  scale_colour_manual(values = setNames(df_funnel$text_col, df_funnel$step), guide = "none") +

  geom_text(
    aes(label = paste0(round(pct_of_total, 1), "%")),
    hjust  = -0.25,
    colour = "#2C2C2A",
    size   = 3.2
  ) +

  coord_flip() +

  scale_y_continuous(
    expand = expansion(mult = c(0, 0.28)),
    labels = comma
  ) +

  labs(
    title    = "ECD Connect: Onboarding funnel",
    subtitle = "Active users registered April 2025 onwards | OA & WL tenants",
    x        = NULL,
    y        = "Number of users"
  ) +

  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold", size = 14, colour = "#2C2C2A"),
    plot.subtitle      = element_text(size = 10,     colour = "#5F5E5A", margin = margin(b = 12)),
    plot.caption       = element_text(size = 8,      colour = "#888780", hjust = 0),
    axis.text.y        = element_text(size = 10,     colour = "#2C2C2A"),
    axis.text.x        = element_text(size = 9,      colour = "#888780"),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(colour = "#D3D1C7", linewidth = 0.4),
    panel.grid.minor   = element_blank(),
    plot.margin        = margin(16, 24, 16, 16)
  )
```

### Reading the funnel

The funnel reveals a two-stage dropout pattern that shapes the entire product strategy:

**Stage 1: Identity friction (Steps 1-2).** 35% of users never complete basic profile steps. This likely reflects a mix of speculative registration via WhatsApp outreach, users who don't have documentation available at the time of signing up, and non-ECD-practitioners who joined to explore the platform. User support data suggests a meaningful subset are prospective practitioners looking for information rather than active users ready to onboard.

**Stage 2: Structural barriers (Steps 3-5).** Connecting to a preschool and adding children requires institutional access. A practitioner who wants to use the app but isn't the registered principal is effectively blocked. In addition, practitioners joining mid-year at larger schools also flag child registration as a friction point: adding a full class roster manually is a significant upfront time investment.

69% of users were last seen on the same day they registered, showing that the funnel doesn't lose people gradually, it loses them immediately.

<div class="insight-box green">
<p class="insight-label">Product signal</p>
<p>Users who reach Step 3 convert at significantly higher rates. The biggest lever for funnel completion is shortening the gap between registration and first genuine usefulness, even in a limited trial mode.</p>
</div>
<div class="insight-box yellow">
<p class="insight-label">Design implication</p>
<p>Steps 3 and 4 occasionally invert: some users create a class before formally connecting to a preschool. The data suggests some users naturally explore class creation before completing preschool connection. Making this pathway more visible in the onboarding flow could encourage more users to reach that point rather than dropping off earlier.</p>
</div>

---

## Part 2: Feature engagement

Being "converted" (having a valid preschool connection) is a proxy for meaningful adoption. The analysis below compares feature engagement rates between all registered users and this converted cohort, revealing which features are true retention drivers versus which are surface-level signals.

```{r feature-engagement}
# ── Load data ─────────────────────────────────────────────────────────────────
# Expected CSV columns: feature, pct_of_all, pct_of_converted
df_features <- read_csv("~/Desktop/data/data-usage/features.csv")

# ── Reshape ───────────────────────────────────────────────────────────────────
df_long <- df_features |>
  select(feature, pct_of_all, pct_of_converted) |>
  mutate(feature = fct_reorder(feature, pct_of_converted)) |>
  pivot_longer(
    cols      = c(pct_of_all, pct_of_converted),
    names_to  = "group",
    values_to = "pct"
  ) |>
  mutate(
    group = case_when(
      group == "pct_of_all"       ~ "All registered users",
      group == "pct_of_converted" ~ "Converted users only"
    ),
    group = fct_relevel(group, "All registered users", "Converted users only")
  )

# ── Palette ───────────────────────────────────────────────────────────────────
colours <- c(
  "All registered users" = "#B4B2A9",
  "Converted users only" = "#1D9E75"
)

# ── Plot ──────────────────────────────────────────────────────────────────────
ggplot(df_long, aes(x = feature, y = pct, fill = group)) +

  geom_col(position = position_dodge(width = 0.7), width = 0.6) +

  geom_text(
    aes(label = paste0(pct, "%")),
    position = position_dodge(width = 0.7),
    hjust    = -0.15,
    size     = 3.0,
    colour   = "#444441"
  ) +

  coord_flip() +

  scale_fill_manual(values = colours, name = NULL) +

  scale_y_continuous(
    expand = expansion(mult = c(0, 0.25)),
    labels = function(x) paste0(x, "%")
  ) +

  labs(
    title    = "ECD Connect: Feature engagement",
    subtitle = "% of users who have ever used each feature · lifetime · OA & WL tenants",
    x        = NULL,
    y        = "% of users",
    caption  = paste0(
      "All registered users n = 3,276 · Converted users (valid preschool connection) n = 1,263\n",
      "Converted users have completed onboarding; all others include trial and registered-only users"
    )
  ) +

  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 14, colour = "#2C2C2A"),
    plot.subtitle    = element_text(size = 10,     colour = "#5F5E5A", margin = margin(b = 12)),
    plot.caption     = element_text(size = 8,      colour = "#888780", hjust = 0),
    axis.text.y      = element_text(size = 10,     colour = "#2C2C2A"),
    axis.text.x      = element_text(size = 9,      colour = "#888780"),
    legend.position  = "top",
    legend.text      = element_text(size = 10,     colour = "#2C2C2A"),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(colour = "#D3D1C7", linewidth = 0.4),
    panel.grid.minor   = element_blank(),
    plot.margin      = margin(16, 24, 16, 16)
  )
```

### Reading the feature data

The gap between grey (all users) and green (converted users) bars is the key signal. Large gaps indicate depth features used after commitment is established. Small gaps indicate hook features used regardless of onboarding status.

<div class="insight-box teal">
<p class="insight-label">Strongest hook feature</p>
<p>Community Profile is the most-used feature overall and shows meaningful uptake even among non-converted users (15.6% vs 24.3%). This suggests practitioners are finding value in the community layer before completing onboarding — a strong signal for surfacing it earlier in the user journey.</p>
</div>

<div class="insight-box green">
<p class="insight-label">Core retention drivers</p>
<p>Activity Planning, Add Child, and Attendance show the largest gaps between all users and converted users, confirming these are depth features used by practitioners who have committed to the platform. Activity Planning in particular sits surprisingly high given it requires lesson planning behaviour change.</p>
</div>

<div class="insight-box pink">
<p class="insight-label">Watch list</p>
<p>Finance tools and Progress Reports show low adoption across both groups. Finance likely reflects a trust barrier — practitioners need confidence in the platform before logging sensitive financial data. Training and Self Assessment also show low numbers, though both features were released significantly later than others so direct comparison is limited.</p>
</div>

---

## Part 3: The retention cliff

### What the data shows

69% of registered users were last seen on the same day they registered. The remaining 31% return at least once, and among those who do, a consistent pattern emerges: the longer a user stays active, the more likely they are to have converted. Conversion rate rises from 31% among same-day users to 95% among users active for 180 or more days.

```{r cliff}
df_cliff <- read_csv("~/Desktop/data/data-usage/retention-cliff.csv")

df_cliff <- df_cliff |>
  mutate(
    days_active_bucket = factor(days_active_bucket, levels = c(
      "0: Same day only", "1-2 days", "3-6 days",
      "7-29 days", "30-89 days", "90-179 days", "180+ days"
    )),
    conversion_status = factor(conversion_status,
                               levels = c("Not converted", "Converted"))
  )

ggplot(df_cliff, aes(x = days_active_bucket, y = users, fill = conversion_status)) +
  geom_col(position = "stack", width = 0.65) +
  scale_fill_manual(
    values = c("Not converted" = "#B4B2A9", "Converted" = "#1DBADF"),
    name = NULL
  ) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.08))) +
  labs(
    title    = "ECD Connect: Retention and conversion by active window",
    subtitle = "Users grouped by days between registration and last login | OA & WL tenants",
    x        = NULL,
    y        = "Number of users",
    caption  = "Deduplicated by phone number. Same-day bucket may include users who completed setup in a single session."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold", size = 14, colour = "#2C2C2A"),
    plot.subtitle      = element_text(size = 10, colour = "#5F5E5A", margin = margin(b = 12)),
    plot.caption       = element_text(size = 8,  colour = "#888780", hjust = 0),
    axis.text.x        = element_text(size = 10, colour = "#2C2C2A"),
    axis.text.y        = element_text(size = 9,  colour = "#888780"),
    legend.position    = "top",
    legend.text        = element_text(size = 10, colour = "#2C2C2A"),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(colour = "#D3D1C7", linewidth = 0.4),
    panel.grid.minor   = element_blank(),
    plot.margin        = margin(16, 24, 16, 16)
  )
```

### Reading the chart

Each bar shows users in that active window, split by conversion status. Conversion rate rises consistently with time in the app: 31% of same-day users converted, rising to 47% at 1-2 days, 58% at 3-6 days, 74% at 30-89 days, and 95% among users active for 180 or more days.

<div class="insight-box teal">
<p class="insight-label">The cliff is real</p>
<p>69% of users never return after registration day. Reminders and nudges are unlikely to move this group. The barrier is structural: users who were never ready to onboard, were blocked by the principal model, or joined out of curiosity rather than immediate intent.</p>
</div>

<div class="insight-box green">
<p class="insight-label">Time and conversion reinforce each other</p>
<p>Users who stay active longer convert at higher rates, and converted users stay active longer. Anything that accelerates time-to-conversion also improves long-term retention. These are not separate problems to solve.</p>
</div>

<div class="insight-box yellow">
<p class="insight-label">Same-day converted users are a distinct group</p>
<p>713 users both converted and were last seen on registration day, likely principals who set up their preschool in a single session and then went quiet. For this group the question is not activation but sustained habit formation.</p>
</div>

<div class="insight-box navy">
<p class="insight-label">Design implication</p>
<p>The priority intervention is not re-engagement campaigns for dormant users. The higher-leverage investment is in the first session: reducing the time between registration and a first moment of genuine value, so more users cross the threshold from curious to committed before they close the app on day one.</p>
</div>

### What same-day users actually did

Among users last seen on registration day, Community Profile was used by 10.4%, far ahead of any other feature. Activity Planning (3.3%) and Resources (2.6%) were the next most common. Attendance and Add Child were used by fewer than 2%.

This is notable because Community Profile is not the most prominent feature on the home screen. Users are actively seeking it out by name, filling in a join form, and still not returning. The community concept draws people in. What happens after they join is not yet compelling enough to bring them back.

<div class="insight-box teal">
<p class="insight-label">Community as an entry point</p>
<p>Community Profile is used by explorers and committed users alike. It is the platform's most accessible hook. The design question for a second version of this product is not whether to build community, but what kind of community only makes sense inside a professional ECD tool, and cannot be replicated by WhatsApp groups that practitioners are already using.</p>
</div>

<div class="insight-box navy">
<p class="insight-label">What WhatsApp cannot do</p>
<p>WhatsApp groups already exist for most of this user base and are active, familiar, and free. A community feature inside a product platform can do things WhatsApp cannot: surface content based on what a practitioner is doing in the app, show verified professional credentials, connect practitioners by district or school size, and provide structured resources rather than a chat scroll. The data suggests demand for connection is real. The design challenge is building something that adds value WhatsApp does not already provide.</p>
</div>

---

## Part 4: Does organisational support change adoption?

ECD Connect operates across two models: an open access version (ECD Connect, available to any practitioner in South Africa) and white-label versions built for specific partner organisations (Ntataise, Khululeka, and True North). White-label users have an organisation behind their onboarding — trained support staff, existing practitioner networks, and in some cases structured rollout programmes. Open access users are entirely self-directed.

This section asks whether that organisational scaffolding makes a measurable difference to conversion and retention.

```{r tenant-conversion}
df_tenants <- read_csv("~/Desktop/data/data-usage/tenants.csv")

df_tenants <- df_tenants |>
  mutate(
    tenant_label = case_when(
      ApplicationName == "ECD Connect" ~ "ECD Connect (OA)",
      TRUE ~ paste0(ApplicationName, " (WL)")
    ),
    tenant_label = fct_reorder(tenant_label, conversion_rate)
  )

ggplot(df_tenants, aes(x = tenant_label, y = conversion_rate, fill = tenant_label)) +
  geom_col(width = 0.6) +
  geom_text(
    aes(label = paste0(conversion_rate, "%")),
    hjust = -0.2,
    size = 3.5,
    colour = "#2C2C2A"
  ) +
  coord_flip() +
  scale_fill_manual(
    values = c(
      "ECD Connect (OA)"       = "#B4B2A9",
      "NtataiseConnect (WL)"   = "#1DBADF",
      "KHULUconnect (WL)"      = "#83BB26",
      "True North Connect (WL)"= "#FFD525"
    ),
    guide = "none"
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.2)),
    labels = function(x) paste0(x, "%")
  ) +
  labs(
    title    = "Conversion rate by tenant",
    subtitle = "% of registered users who completed onboarding to a valid preschool connection",
    x        = NULL,
    y        = "Conversion rate"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold", size = 14, colour = "#2C2C2A"),
    plot.subtitle      = element_text(size = 10, colour = "#5F5E5A", margin = margin(b = 12)),
    axis.text.y        = element_text(size = 10, colour = "#2C2C2A"),
    axis.text.x        = element_text(size = 9,  colour = "#888780"),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(colour = "#D3D1C7", linewidth = 0.4),
    panel.grid.minor   = element_blank(),
    plot.margin        = margin(16, 24, 16, 16)
  )
```

### Reading the data

Two of the three white-label tenants convert at substantially higher rates than open access: Ntataise at 65.3% and Khululeka at 54.5%, compared to 40.8% for ECD Connect OA. Converted users on white-label tenants also stay active significantly longer — Ntataise and Khululeka converted users average over 70 days active, compared to 31 days for OA converted users.

True North is the exception: conversion rate (36.4%) sits below OA, but their converted users are the most active of any tenant at an average of 88 days. With only 33 total users, True North's numbers should be treated cautiously, but the pattern may reflect the dynamics of a smaller, more tightly managed rollout where fewer users convert but those who do are highly committed.

Across all tenants, the gap between converted and non-converted avg days active is stark. Non-converted users barely engage regardless of tenant type, averaging under 5 days across the board.

<div class="insight-box teal">
<p class="insight-label">Organisational support drives conversion</p>
<p>The data is consistent: users with an organisation behind their onboarding convert at higher rates and stay active longer. This is not primarily a product quality finding — the same product produces meaningfully different outcomes depending on the support model around it. For any new deployment of this platform, the onboarding support model is as important a design decision as the product itself.</p>
</div>

<div class="insight-box green">
<p class="insight-label">OA conversion is strong given the context</p>
<p>40.8% conversion among entirely self-directed users with no organisational support is a meaningful baseline. These are practitioners who found the platform independently, registered without prompting, and nearly half completed full onboarding on their own. The OA channel is working; the question is what light-touch support could move that number further.</p>
</div>

<div class="insight-box yellow">
<p class="insight-label">Room to grow with coaching features</p>
<p>None of the partner organisations have yet activated the platform's coaching feature, which enables structured support between coordinators and practitioners. Khululeka has plans to pilot this. If coaching drives the kind of engagement uplift seen in the broader WL vs OA comparison, it could be a meaningful lever for both conversion and long-term retention.</p>
</div>

<div class="insight-box navy">
<p class="insight-label">Implication for SmartStart</p>
<p>A self-directed open access product will always face a conversion ceiling. The data suggests that ceiling sits somewhere around 40%. Pushing meaningfully beyond that requires organisational infrastructure: trained support staff, structured onboarding programmes, and principal-level engagement. Building a great product is necessary but not sufficient for adoption at scale in this context.</p>
</div>

---

## Part 5: Principals vs practitioners

Of the 4,021 registered users, 1,473 are identifiable as principals, 182 as practitioners, and 2,366 remain unclassified — users who registered but never completed enough onboarding to take on either role. The 8:1 ratio of principals to practitioners is itself a finding: the platform is currently used almost entirely by principals, and practitioner adoption is entirely downstream of principal engagement.

```{r role-features}
df_roles <- read_csv("~/Desktop/data/data-usage/role-features.csv")

df_roles <- df_roles |>
  mutate(
    Role = factor(Role, levels = c("Principal", "Practitioner")),
    feature = fct_reorder(feature, pct_of_role, .fun = max)
  )

df_roles <- df_roles |>
  filter(feature != "Income Statements")

ggplot(df_roles, aes(x = feature, y = pct_of_role, fill = Role)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(
    aes(label = paste0(pct_of_role, "%")),
    position = position_dodge(width = 0.7),
    hjust = -0.15,
    size = 3.0,
    colour = "#444441"
  ) +
  coord_flip() +
  scale_fill_manual(
    values = c("Principal" = "#1DBADF", "Practitioner" = "#83BB26"),
    name = NULL
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.25)),
    labels = function(x) paste0(x, "%")
  ) +
  labs(
    title    = "ECD Connect: Feature usage by role",
    subtitle = "% of principals and practitioners who have ever used each feature",
    x        = NULL,
    y        = "% of users",
    caption  = "Principals n = 1,473 · Practitioners n = 182 · Excludes unclassified users"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold", size = 14, colour = "#2C2C2A"),
    plot.subtitle      = element_text(size = 10, colour = "#5F5E5A", margin = margin(b = 12)),
    plot.caption       = element_text(size = 8,  colour = "#888780", hjust = 0),
    axis.text.y        = element_text(size = 10, colour = "#2C2C2A"),
    axis.text.x        = element_text(size = 9,  colour = "#888780"),
    legend.position    = "top",
    legend.text        = element_text(size = 10, colour = "#2C2C2A"),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(colour = "#D3D1C7", linewidth = 0.4),
    panel.grid.minor   = element_blank(),
    plot.margin        = margin(16, 24, 16, 16)
  )
```

### Reading the data

The chart shows feature usage rates within each role. Practitioners show higher rates on classroom-facing features; principals dominate on administrative ones.

<div class="insight-box teal">
<p class="insight-label">The platform is principal-led</p>
<p>With an 8:1 ratio of principals to practitioners, the platform currently functions primarily as a tool for school leaders rather than classroom teachers. Practitioner access depends entirely on principal setup, meaning principal recruitment and onboarding quality is the single most important variable in overall platform adoption.</p>
</div>

<div class="insight-box green">
<p class="insight-label">Practitioners are the classroom users</p>
<p>Once onboarded, practitioners use every classroom-facing feature at significantly higher rates than principals. Attendance is used by 40.6% of practitioners vs 10.2% of principals. Calendar (29.4% vs 9.4%) and Progress Reports (17.5% vs 5%) show the same pattern. Community Profile is the one exception, used at nearly identical rates by both roles (26.3% vs 25.9%), suggesting it serves a different need — professional connection rather than classroom management.</p>
</div>

<div class="insight-box pink">
<p class="insight-label">Principals do the admin, practitioners do the classroom work</p>
<p>Income Statements is used by 12% of principals and virtually no practitioners, confirming the expected role split on administrative features. The data suggests the platform has naturally divided into two distinct use cases: principals managing the school, practitioners managing the classroom. A new platform build should design these two experiences more deliberately rather than treating all users as a single audience.</p>
</div>

<div class="insight-box navy">
<p class="insight-label">Implication for SmartStart</p>
<p>Designing around the principal workflow is the right starting point: principals are the gatekeepers, the multipliers, and currently the primary users. But practitioners who do get access engage deeply with classroom features. A new platform should make it as easy as possible for principals to bring practitioners on, then design the practitioner experience around daily classroom workflows.</p>
</div>

---

## Part 6: Feature sequencing

Among converted users who used at least one feature, what did they do first? This matters for onboarding design: if there is a dominant entry point, that feature should be prominent in the first session experience.

The analysis covers all converted users with at least one trackable feature event. Users whose first action was creating a class (a required onboarding step rather than a discretionary feature) are excluded so the chart reflects genuine feature choices.

```{r feature-sequencing}
df_seq <- read_csv("~/Desktop/data/data-usage/feature-sequencing.csv")

df_seq <- df_seq |>
  filter(!first_feature %in% c("Created Class", "Added Practitioner")) |>
  mutate(first_feature = fct_reorder(first_feature, users))

ggplot(df_seq, aes(x = first_feature, y = pct_of_converted)) +
  geom_col(fill = "#1DBADF", width = 0.65) +
  geom_text(
    aes(label = paste0(pct_of_converted, "%")),
    hjust = -0.2,
    size = 3.2,
    colour = "#2C2C2A"
  ) +
  coord_flip() +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.2)),
    labels = function(x) paste0(x, "%")
  ) +
  labs(
    title    = "ECD Connect: First feature used by converted users",
    subtitle = "% of converted users whose first discretionary feature use was each feature",
    x        = NULL,
    y        = "% of converted users",
    caption  = "Excludes class creation (required onboarding step). Converted users n = 1,670."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold", size = 14, colour = "#2C2C2A"),
    plot.subtitle      = element_text(size = 10, colour = "#5F5E5A", margin = margin(b = 12)),
    plot.caption       = element_text(size = 8,  colour = "#888780", hjust = 0),
    axis.text.y        = element_text(size = 10, colour = "#2C2C2A"),
    axis.text.x        = element_text(size = 9,  colour = "#888780"),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(colour = "#D3D1C7", linewidth = 0.4),
    panel.grid.minor   = element_blank(),
    plot.margin        = margin(16, 24, 16, 16)
  )
```

### Reading the chart

Community Profile is the first discretionary feature for 16.8% of converted users, nearly double the next feature. Attendance (7.8%) and Add Child (6.9%) follow, reflecting the core administrative workflow. Activity Planning (6.2%) shows a meaningful group drawn in by the content and curriculum layer before anything else. Everything else sits below 4%.

Note that 54% of converted users have no trackable feature events at all, meaning they completed preschool setup but never used a tracked feature. These are likely principals who set up the preschool structure and then stepped back, or users who completed setup in a single session and did not return.

<div class="insight-box teal">
<p class="insight-label">Community Profile is the dominant entry point</p>
<p>Among users who go beyond setup, Community Profile is the most common first feature by a significant margin. Combined with its strong showing in the overall feature engagement data and among same-day users, this is consistent evidence that community is the platform's primary hook. It draws in explorers, it draws in converted users, and it is used equally by principals and practitioners. Whatever form community takes in a new platform build, it should be prominent from day one.</p>
</div>

<div class="insight-box green">
<p class="insight-label">Two distinct entry paths</p>
<p>The sequencing data suggests two types of converted user. The first goes straight to administrative setup: Add Child, Attendance, Income Statements. The second explores content and community first: Community Profile, Activity Planning, Resources. These are different user intents and may benefit from different onboarding paths rather than a single linear flow.</p>
</div>

<div class="insight-box navy">
<p class="insight-label">Implication for SmartStart</p>
<p>A new platform should not assume a single onboarding path. Principals coming in with clear administrative intent need a fast route to preschool setup and child registration. Users coming in to explore need an easy on-ramp to community and content features before they hit the administrative setup wall. Designing for both intents from the start would likely improve conversion rates for both groups.</p>
</div>

---


## Methodology note

All data is aggregate and anonymised. No individual practitioner or child data is reported. Funnel data covers users registered from October 2024 with activity on OA (open access) and WL (white-label) tenants. Feature engagement reflects lifetime usage through the analysis date. Conversion is defined as having at least one valid preschool connection established.

---

*Analysis by Kim Tichmann · Data & Product Strategy Lead, ECD Connect*

*ECD Connect is a product of [DGMT](https://dgmt.co.za) (DG Murray Trust). All views are those of the author.*
